ECCCos from the Black Box:\\ Faithful Explanations through\\ Energy-Constrained Conformal Counterfactuals

### 2023-05-10 notebooks/fashion_mnist.qmd
paper

```{julia}
# Measures:
measures = [
    CounterfactualExplanations.distance,
    ECCCo.distance_from_energy,
    ECCCo.distance_from_targets,
    CounterfactualExplanations.Evaluation.validity,
    CounterfactualExplanations.Evaluation.redundancy,
]

>>>>

```{julia}
# Measures:
measures = [
    CounterfactualExplanations.distance,
    ECCCo.distance_from_energy,
    ECCCo.distance_from_targets,
    CounterfactualExplanations.Evaluation.validity,
    CounterfactualExplanations.Evaluation.redundancy,
    ECCCo.set_size_penalty
]

### 2023-05-10 notebooks/mnist.qmd
paper

```{julia}
# Measures:
measures = [
    CounterfactualExplanations.distance,
    ECCCo.distance_from_energy,
    ECCCo.distance_from_targets,
    CounterfactualExplanations.Evaluation.validity,
    CounterfactualExplanations.Evaluation.redundancy,
]

>>>>

```{julia}
# Measures:
measures = [
    CounterfactualExplanations.distance,
    ECCCo.distance_from_energy,
    ECCCo.distance_from_targets,
    CounterfactualExplanations.Evaluation.validity,
    CounterfactualExplanations.Evaluation.redundancy,
    ECCCo.set_size_penalty,
]

### 2023-05-10 notebooks/tables.Rmd
paper

```{r}
measures <- c("distance_from_energy", "distance_from_targets")
tab <- dt[
    variable %in% measures & source=="real",
    .(
      value=sprintf("%1.2f (%1.2f)", mean(value), sd(value)),
      val = mean(value)
    ),
    .(dataname, generator, model, variable)
]
tab[,min_val:=val==min(val),.(model, dataname, variable)]
tab$value <- cell_spec(tab$value, "latex", bold=tab$min_val)
tab[,val:=NULL]
tab[,min_val:=NULL]
tab[generator == "eccco", generator:="ECCCo"]
tab[generator == "revise", generator:="REVISE"]
tab[generator == "generic", generator:="Wachter"]
tab[generator == "wachter", generator:="Wachter"]
tab[generator == "greedy", generator:="Schut"]
n_data <- length(unique(tab$dataname))
tab <- dcast(tab, model + generator ~ dataname + variable)
col_names <- c(
  "Model",
  "Generator",
  rep(c("Non-conformity ↓","Implausibility ↓"),n_data)
)
kbl(
  tab, caption = "Results for real-world datasets. Standard deviation in parentheses. \\label{tab:results-real}", 
  align = "c", col.names=col_names, booktabs = T, escape=F, 
  format="latex", linesep = c("", "", "", "\\addlinespace") 
) %>%
  kable_styling(latex_options = c("scale_down")) %>%
  add_header_above(c(" " = 2, "California Housing" = 2, "GMSC" = 2, "MNIST" = 2)) %>%
  column_spec(1, bold = T) %>%
  collapse_rows(columns = 1:2, latex_hline = "major", valign = "middle") %>%
  save_kable("paper/contents/table_real.tex")
```

>>>>

## Main tables

```{r}
measures <- c("distance_from_energy", "distance_from_targets")
tab <- dt[
    variable %in% measures & source=="real",
    .(
      value=sprintf("%1.2f (%1.2f)", mean(value), sd(value)),
      val = mean(value),
      std = sd(value)
    ),
    .(dataname, generator, model, variable)
]
tab[,min_val:=val==min(val),.(model, dataname, variable)]
tab[,two_std_wachter:=val+2*std<val[generator=="Wachter"],.(model, dataname, variable)]
tab[,one_std_wachter:=val+1*std<val[generator=="Wachter"],.(model, dataname, variable)]
tab[one_std_wachter==T,value:=paste0(value,"*")]
tab[two_std_wachter==T,value:=paste0(value,"*")]
tab$value <- cell_spec(tab$value, "latex", bold=tab$min_val)
tab[,val:=NULL]
tab[,std:=NULL]
tab[,min_val:=NULL]
tab[,two_std_wachter:=NULL]
tab[,one_std_wachter:=NULL]
n_data <- length(unique(tab$dataname))
tab <- dcast(tab, model + generator ~ dataname + variable)
col_names <- c(
  "Model",
  "Generator",
  rep(c("Non-conformity ↓","Implausibility ↓"),n_data)
)
kbl(
  tab, caption = "Results for real-world datasets. Standard deviations across samples are shown in parentheses. Best outcomes are highlighted in bold. Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter). \\label{tab:results-real} \\newline", 
  align = "c", col.names=col_names, booktabs = T, escape=F, 
  format="latex", linesep = c("", "", "", "\\addlinespace") 
) %>%
  kable_styling(latex_options = c("scale_down")) %>%
  kable_paper(full_width = F) %>%
  add_header_above(c(" " = 2, "California Housing" = 2, "GMSC" = 2, "MNIST" = 2)) %>%
  column_spec(1, bold = T) %>%
  collapse_rows(columns = 1:2, latex_hline = "major", valign = "middle") %>%
  save_kable("paper/contents/table_real.tex")
```

### 2023-05-10 paper/paper.tex
paper

% Numbered Environments:
\newtheorem{definition}{Definition}[section]

@@

\subsection{From Fidelity to Model Conformity}

@@

\begin{definition}[Conformal Counterfactuals]
  \label{def:conformal}
  Let $\mathcal{X}_{\theta}|\mathbf{y}^* = p_{\theta}(x|\mathbf{y}^*)$ denote the conditional distribution of $\mathbf{x}$ in the target class $\mathbf{y}^*$, where $\theta$ denotes the parameters of model $M_{\theta}$. Then for $\mathbf{x}^{\prime}$ to be considered a conformal counterfactual, we need: $\mathbf{x}^{\prime} \sim \mathcal{X}_{\theta}|\mathbf{y}^*$.
\end{definition}

In words, conformal counterfactuals conform with what the predictive model has learned about the input data $\mathbf{x}$. Since this definition works with distributional properties, it explicitly accounts for the multiplicity of explanations we discussed earlier. To assess counterfactuals with respect to Definition~\ref{def:conformal}, we need to be able to quantify the posterior conditional distribution $p_{\theta}(\mathbf{x}|\mathbf{y}^*)$. This is very much at the core of our proposed methodological framework, which reconciles the notions of plausibility and model conformity and which we will introduce next.

@@

where $\mathbf{r}_j \sim \mathcal{N}(\mathbf{0},\mathbf{I})$ is the stochastic term and the step-size $\epsilon$ is typically polynomially decayed. The term $\mathcal{E}(\mathbf{x}_j|\mathbf{y}^*)$ denotes the energy function where we use $\mathcal{E}(\mathbf{x}_j|\mathbf{y}^*)=-M_{\theta}(\mathbf{x}_j)[\mathbf{y}^*]$, that is the negative logit corresponding to the target class label $\mathbf{y}^*$. Generating multiple samples in this manner yields an empirical distribution $\hat{\mathcal{X}}_{\theta}|\mathbf{y}^*$ that we use in our search for plausible counterfactuals, as discussed in more detail below. Appendix~\ref{app-jem} provides additional implementation details for any tasks related to energy-based modelling.

@@

The first two terms in Equation~\ref{eq:eccco} correspond to the counterfactual search objective defined in~\citet{wachter2017counterfactual} which merely penalises the distance of counterfactuals from their factual values. The additional two penalties in ECCCo ensure that counterfactuals conform with the model's generative property and lead to minimally uncertain predictions, respectively. The hyperparameters $\lambda_1, ..., \lambda_3$ can be used to balance the different objectives: for example, we may choose to incur larger deviations from the factual in favour of conformity with the model's generative property by choosing lower values of $\lambda_1$ and relatively higher values of $\lambda_2$. Figure~\ref{fig:eccco} illustrates this balancing act for an example involving synthetic data: vector fields indicate the direction of gradients with respect to the different components our proposed objective function (Equation~\ref{eq:eccco}).

@@

In this section, we bolster our anecdotal findings from the previous section through rigorous empirical analysis. We first briefly describe our evaluation framework and data, before presenting and discussing our results.

\subsection{Evaluation Measures}\label{evaluation}

Above we have defined plausibility (Definition~\ref{def:plausible}) and conformity (Definition~\ref{def:conformal}) for Counterfactual Explanations. In this subsection, we introduce evaluation measures that facilitate a quantitative evaluation of counterfactuals for these objectives.

Firstly, in order to assess the plausibility of counterfactuals we adapt the implausibility metric proposed in~\citet{guidotti2022counterfactual}. The authors propose to evaluate plausibility in terms of the distance of the counterfactual $\mathbf{x}^{\prime}$ from its nearest neighbour in the target class $\mathbf{y}^*$: the smaller this distance, the more plausible the counterfactual. Instead of focusing only on the nearest neighbour of $\mathbf{x}^{\prime}$, we suggest computing the average over distances from multiple (possibly all) observed instances in the target class. Formally, for a single counterfactual, we have:

\begin{equation}\label{eq:impl}
  \begin{aligned}
    \text{impl} = \frac{1}{\lvert\mathbf{x} \in \mathcal{X}|\mathbf{y}^*\rvert} \sum_{\mathbf{x} \in \mathcal{X}|\mathbf{y}^*} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
  \end{aligned}
\end{equation}

This measure is straightforward to compute and should be less sensitive to outliers in the target class than the one based on the nearest neighbour. It also gives rise to a very similar evaluation measure for conformity. We merely swap out the subsample of individuals in the target class for the empirical distribution of generated conditional samples:

\begin{equation}\label{eq:conf}
  \begin{aligned}
    \text{conf} = \frac{1}{\lvert\mathbf{x} \in \mathcal{X}_{\theta}|\mathbf{y}^*\rvert} \sum_{\mathbf{x} \in \mathcal{X}_{\theta}|\mathbf{y}^*} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
  \end{aligned}
\end{equation}

As noted by~\citet{guidotti2022counterfactual}, these distance-based measures are simplistic and more complex alternative measures may ultimately be more appropriate for the task. For example, we considered using statistical divergence measures instead. This would involve generating not one but many counterfactuals and comparing the generated empirical distribution to the target distributions in Definitions~\ref{def:plausible} and~\ref{def:conformal}. While this approach is potentially more rigorous, generating enough counterfactuals is not always practical.

@@

\begin{itemize}
  \item BatchNorm does not seem compatible with JEM
  \item Coverage and temperature impacts CCE in somewhat unpredictable ways
  \item It seems that models that are not explicitly trained for generative task, still learn it implictly
  \item Batch size seems to impact quality of generated samples (at inference, but not so much during JEM training)
  \item SGLD takes time 
  \item REVISE has benefit of lower dimensional space
  \item For MNIST it seems that ECCCo is better at reducing pixel values than increasing them (better at erasing than writing)
  \item JEMs are more difficult to train
  \item There is a tradeoff: higher cost vs. higher conformity/plausibility
  \item Results are sensitive to choices of penalty strength and step size
  \item Counterfactuals may end up looking fairly homogenous
  \item For MNIST data we found CP to have little effect
  \item JEMs themselves are sensitive to scale
  \item ECCCo can backfire, in case generative property of model is poor
\end{itemize}

@@

\section{Submission of papers to NeurIPS 2023}

Please read the instructions below carefully and follow them faithfully.

\subsection{Style}

Papers to be submitted to NeurIPS 2023 must be prepared according to the
instructions presented here. Papers may only be up to {\bf nine} pages long,
including figures. Additional pages \emph{containing only acknowledgments and
references} are allowed. Papers that exceed the page limit will not be
reviewed, or in any other way considered for presentation at the conference.

The margins in 2023 are the same as those in previous years.

Authors are required to use the NeurIPS \LaTeX{} style files obtainable at the
NeurIPS website as indicated below. Please make sure you use the current files
and not previous versions. Tweaking the style files may be grounds for
rejection.

\subsection{Retrieval of style files}

The style files for NeurIPS and other conference information are available on
the website at
\begin{center}
  \url{http://www.neurips.cc/}
\end{center}
The file \verb+neurips_2023.pdf+ contains these instructions and illustrates the
various formatting requirements your NeurIPS paper must satisfy.

The only supported style file for NeurIPS 2023 is \verb+neurips_2023.sty+,
rewritten for \LaTeXe{}.  \textbf{Previous style files for \LaTeX{} 2.09,
  Microsoft Word, and RTF are no longer supported!}

The \LaTeX{} style file contains three optional arguments: \verb+final+, which
creates a camera-ready copy, \verb+preprint+, which creates a preprint for
submission to, e.g., arXiv, and \verb+nonatbib+, which will not load the
\verb+natbib+ package for you in case of package clash.

\paragraph{Preprint option}
If you wish to post a preprint of your work online, e.g., on arXiv, using the
NeurIPS style, please use the \verb+preprint+ option. This will create a
nonanonymized version of your work with the text ``Preprint. Work in progress.''
in the footer. This version may be distributed as you see fit, as long as you do not say which conference it was submitted to. Please \textbf{do
  not} use the \verb+final+ option, which should \textbf{only} be used for
papers accepted to NeurIPS.

At submission time, please omit the \verb+final+ and \verb+preprint+
options. This will anonymize your submission and add line numbers to aid
review. Please do \emph{not} refer to these line numbers in your paper as they
will be removed during generation of camera-ready copies.

The file \verb+neurips_2023.tex+ may be used as a ``shell'' for writing your
paper. All you have to do is replace the author, title, abstract, and text of
the paper with your own.

The formatting instructions contained in these style files are summarized in
Sections \ref{gen_inst}, \ref{headings}, and \ref{others} below.

\section{General formatting instructions}
\label{gen_inst}

The text must be confined within a rectangle 5.5~inches (33~picas) wide and
9~inches (54~picas) long. The left margin is 1.5~inch (9~picas).  Use 10~point
type with a vertical spacing (leading) of 11~points.  Times New Roman is the
preferred typeface throughout, and will be selected for you by default.
Paragraphs are separated by \nicefrac{1}{2}~line space (5.5 points), with no
indentation.

The paper title should be 17~point, initial caps/lower case, bold, centered
between two horizontal rules. The top rule should be 4~points thick and the
bottom rule should be 1~point thick. Allow \nicefrac{1}{4}~inch space above and
below the title to rules. All pages should start at 1~inch (6~picas) from the
top of the page.

For the final version, authors' names are set in boldface, and each name is
centered above the corresponding address. The lead author's name is to be listed
first (left-most), and the co-authors' names (if different address) are set to
follow. If there is only one co-author, list both author and co-author side by
side.

Please pay special attention to the instructions in Section \ref{others}
regarding figures, tables, acknowledgments, and references.

\section{Headings: first level}
\label{headings}

All headings should be lower case (except for first word and proper nouns),
flush left, and bold.

First-level headings should be in 12-point type.

\subsection{Headings: second level}

Second-level headings should be in 10-point type.

\subsubsection{Headings: third level}

Third-level headings should be in 10-point type.

\paragraph{Paragraphs}

There is also a \verb+\paragraph+ command available, which sets the heading in
bold, flush left, and inline with the text, with the heading followed by 1\,em
of space.

\section{Citations, figures, tables, references}
\label{others}

These instructions apply to everyone.

\subsection{Citations within the text}

The \verb+natbib+ package will be loaded for you by default.  Citations may be
author/year or numeric, as long as you maintain internal consistency.  As to the
format of the references themselves, any style is acceptable as long as it is
used consistently.

The documentation for \verb+natbib+ may be found at
\begin{center}
  \url{http://mirrors.ctan.org/macros/latex/contrib/natbib/natnotes.pdf}
\end{center}
Of note is the command \verb+\citet+, which produces citations appropriate for
use in inline text.  For example,
\begin{verbatim}
   \citet{hasselmo} investigated\dots
\end{verbatim}
produces
\begin{quote}
  Hasselmo, et al.\ (1995) investigated\dots
\end{quote}

If you wish to load the \verb+natbib+ package with options, you may add the
following before loading the \verb+neurips_2023+ package:
\begin{verbatim}
   \PassOptionsToPackage{options}{natbib}
\end{verbatim}

If \verb+natbib+ clashes with another package you load, you can add the optional
argument \verb+nonatbib+ when loading the style file:
\begin{verbatim}
   \usepackage[nonatbib]{neurips_2023}
\end{verbatim}

As submission is double blind, refer to your own published work in the third
person. That is, use ``In the previous work of Jones et al.\ [4],'' not ``In our
previous work [4].'' If you cite your other papers that are not widely available
(e.g., a journal paper under review), use anonymous author names in the
citation, e.g., an author of the form ``A.\ Anonymous'' and include a copy of the anonymized paper in the supplementary material.

\subsection{Footnotes}

Footnotes should be used sparingly.  If you do require a footnote, indicate
footnotes with a number\footnote{Sample of the first footnote.} in the
text. Place the footnotes at the bottom of the page on which they appear.
Precede the footnote with a horizontal rule of 2~inches (12~picas).

Note that footnotes are properly typeset \emph{after} punctuation
marks.\footnote{As in this example.}

\subsection{Figures}

\begin{figure}
  \centering
  \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
  \caption{Sample figure caption.}
\end{figure}

All artwork must be neat, clean, and legible. Lines should be dark enough for
purposes of reproduction. The figure number and caption always appear after the
figure. Place one line space before the figure caption and one line space after
the figure. The figure caption should be lower case (except for first word and
proper nouns); figures are numbered consecutively.

You may use color figures.  However, it is best for the figure captions and the
paper body to be legible if the paper is printed in either black/white or in
color.

\subsection{Tables}

All tables must be centered, neat, clean and legible.  The table number and
title always appear before the table.  See Table~\ref{sample-table}.

Place one line space before the table title, one line space after the
table title, and one line space after the table. The table title must
be lower case (except for first word and proper nouns); tables are
numbered consecutively.

Note that publication-quality tables \emph{do not contain vertical rules.} We
strongly suggest the use of the \verb+booktabs+ package, which allows for
typesetting high-quality, professional tables:
\begin{center}
  \url{https://www.ctan.org/pkg/booktabs}
\end{center}
This package was used to typeset Table~\ref{sample-table}.

\begin{table}
  \caption{Sample table title}
  \label{sample-table}
  \centering
  \begin{tabular}{lll}
    \toprule
    \multicolumn{2}{c}{Part}                   \\
    \cmidrule(r){1-2}
    Name     & Description     & Size ($\mu$m) \\
    \midrule
    Dendrite & Input terminal  & $\sim$100     \\
    Axon     & Output terminal & $\sim$10      \\
    Soma     & Cell body       & up to $10^6$  \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Math}
Note that display math in bare TeX commands will not create correct line numbers for submission. Please use LaTeX (or AMSTeX) commands for unnumbered display math. (You really shouldn't be using \$\$ anyway; see \url{https://tex.stackexchange.com/questions/503/why-is-preferable-to} and \url{https://tex.stackexchange.com/questions/40492/what-are-the-differences-between-align-equation-and-displaymath} for more information.)

\subsection{Final instructions}

Do not change any aspects of the formatting parameters in the style files.  In
particular, do not modify the width or length of the rectangle the text should
fit into, and do not change font sizes (except perhaps in the
\textbf{References} section; see below). Please note that pages should be
numbered.

\section{Preparing PDF files}

Please prepare submission files with paper size ``US Letter,'' and not, for
example, ``A4.''

Fonts were the main cause of problems in the past years. Your PDF file must only
contain Type 1 or Embedded TrueType fonts. Here are a few instructions to
achieve this.

\begin{itemize}

\item You should directly generate PDF files using \verb+pdflatex+.

\item You can check which fonts a PDF files uses.  In Acrobat Reader, select the
  menu Files$>$Document Properties$>$Fonts and select Show All Fonts. You can
  also use the program \verb+pdffonts+ which comes with \verb+xpdf+ and is
  available out-of-the-box on most Linux machines.

\item \verb+xfig+ "patterned" shapes are implemented with bitmap fonts.  Use
  "solid" shapes instead.

\item The \verb+\bbold+ package almost always uses bitmap fonts.  You should use
  the equivalent AMS Fonts:
\begin{verbatim}
   \usepackage{amsfonts}
\end{verbatim}
followed by, e.g., \verb+\mathbb{R}+, \verb+\mathbb{N}+, or \verb+\mathbb{C}+
for $\mathbb{R}$, $\mathbb{N}$ or $\mathbb{C}$.  You can also use the following
workaround for reals, natural and complex:
\begin{verbatim}
   \newcommand{\RR}{I\!\!R} %real numbers
   \newcommand{\Nat}{I\!\!N} %natural numbers
   \newcommand{\CC}{I\!\!\!\!C} %complex numbers
\end{verbatim}
Note that \verb+amsfonts+ is automatically loaded by the \verb+amssymb+ package.

\end{itemize}

If your file contains type 3 fonts or non embedded TrueType fonts, we will ask
you to fix it.

\subsection{Margins in \LaTeX{}}

Most of the margin problems come from figures positioned by hand using
\verb+\special+ or other commands. We suggest using the command
\verb+\includegraphics+ from the \verb+graphicx+ package. Always specify the
figure width as a multiple of the line width as in the example below:
\begin{verbatim}
   \usepackage[pdftex]{graphicx} ...
   \includegraphics[width=0.8\linewidth]{myfile.pdf}
\end{verbatim}
See Section 4.4 in the graphics bundle documentation
(\url{http://mirrors.ctan.org/macros/latex/required/graphics/grfguide.pdf})

A number of width problems arise when \LaTeX{} cannot properly hyphenate a
line. Please give LaTeX hyphenation hints using the \verb+\-+ command when
necessary.

@@

\section{Supplementary Material}

Authors may wish to optionally include extra information (complete proofs, additional experiments and plots) in the appendix. All such materials should be part of the supplemental material (submitted separately) and should NOT be included in the main submission.

\section*{References}

References follow the acknowledgments in the camera-ready paper. Use unnumbered first-level heading for
the references. Any choice of citation style is acceptable as long as you are
consistent. It is permissible to reduce the font size to \verb+small+ (9 point)
when listing the references.
Note that the Reference section does not count towards the page limit.
\medskip

{
\small

[1] Alexander, J.A.\ \& Mozer, M.C.\ (1995) Template-based algorithms for
connectionist rule extraction. In G.\ Tesauro, D.S.\ Touretzky and T.K.\ Leen
(eds.), {\it Advances in Neural Information Processing Systems 7},
pp.\ 609--616. Cambridge, MA: MIT Press.

[2] Bower, J.M.\ \& Beeman, D.\ (1995) {\it The Book of GENESIS: Exploring
  Realistic Neural Models with the GEneral NEural SImulation System.}  New York:
TELOS/Springer--Verlag.

[3] Hasselmo, M.E., Schnell, E.\ \& Barkai, E.\ (1995) Dynamics of learning and
recall at excitatory recurrent synapses and cholinergic modulation in rat
hippocampal region CA3. {\it Journal of Neuroscience} {\bf 15}(7):5249-5262.
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

% Numbered Environments:
\newtheorem{definition}{Definition}[section]
\newtheorem{question}{Research Question}[section]

@@

\subsection{From Fidelity to Model Faithfulness}

@@

\begin{definition}[Faithful Counterfactuals]
  \label{def:faithful}
  Let $\mathcal{X}_{\theta}|\mathbf{y}^* = p_{\theta}(\mathbf{X}_{\mathbf{y}^*})$ denote the conditional distribution of $\mathbf{x}$ in the target class $\mathbf{y}^*$, where $\theta$ denotes the parameters of model $M_{\theta}$. Then for $\mathbf{x}^{\prime}$ to be considered a conformal counterfactual, we need: $\mathbf{x}^{\prime} \sim \mathcal{X}_{\theta}|\mathbf{y}^*$.
\end{definition}

In words, conformal counterfactuals conform with what the predictive model has learned about the input data $\mathbf{x}$. Since this definition works with distributional properties, it explicitly accounts for the multiplicity of explanations we discussed earlier. To assess counterfactuals with respect to Definition~\ref{def:faithful}, we need to be able to quantify the posterior conditional distribution $p_{\theta}(\mathbf{x}|\mathbf{y}^*)$. This is very much at the core of our proposed methodological framework, which reconciles the notions of plausibility and model faithfulness and which we will introduce next.

@@

where $\mathbf{r}_j \sim \mathcal{N}(\mathbf{0},\mathbf{I})$ is the stochastic term and the step-size $\epsilon$ is typically polynomially decayed. The term $\mathcal{E}(\mathbf{x}_j|\mathbf{y}^*)$ denotes the energy function where we use $\mathcal{E}(\mathbf{x}_j|\mathbf{y}^*)=-M_{\theta}(\mathbf{x}_j)[\mathbf{y}^*]$, that is the negative logit corresponding to the target class label $\mathbf{y}^*$. Generating multiple samples in this manner yields an empirical distribution $\hat{\mathbf{X}}_{\theta,\mathbf{y}^*}$ that we use in our search for plausible counterfactuals, as discussed in more detail below. Appendix~\ref{app-jem} provides additional implementation details for any tasks related to energy-based modelling.

@@

The first two terms in Equation~\ref{eq:eccco} correspond to the counterfactual search objective defined in~\citet{wachter2017counterfactual} which merely penalises the distance of counterfactuals from their factual values. The additional two penalties in ECCCo ensure that counterfactuals conform with the model's generative property and lead to minimally uncertain predictions, respectively. The hyperparameters $\lambda_1, ..., \lambda_3$ can be used to balance the different objectives: for example, we may choose to incur larger deviations from the factual in favour of faithfulness with the model's generative property by choosing lower values of $\lambda_1$ and relatively higher values of $\lambda_2$. Figure~\ref{fig:eccco} illustrates this balancing act for an example involving synthetic data: vector fields indicate the direction of gradients with respect to the different components our proposed objective function (Equation~\ref{eq:eccco}).

@@

In this section, we present our empirical analysis and findings. Our goal is to shed line on the following questions:

\begin{question}[Feasibility]\label{rq:feasibility}
  Is it feasible to generate plausible Counterfactual Explanations through ECCCo without relying on surrogate models?
\end{question}

\begin{question}[Drivers]\label{rq:drivers}
  Subject to feasibility, what drives the performance of ECCCo? Is it sufficient to rely on energy-based modelling to quantify the model's generative property? Is it sufficient to rely on conformal prediction to quantify the model's uncertainty?
\end{question}

In the following, we first briefly describe our evaluation framework and data, before presenting and discussing our results.

\subsection{Key Evaluation Measures}\label{evaluation}

Above we have defined plausibility (Definition~\ref{def:plausible}) and faithfulness (Definition~\ref{def:faithful}) for Counterfactual Explanations. These are the main criteria we use to evaluate counterfactuals in this study. In order to quantify the plausibility of counterfactuals we use a slightly adapted version of the implausibility metric proposed in~\citet{guidotti2022counterfactual}. Formally, for a single counterfactual, we define implausibility as follows,

\begin{equation}\label{eq:impl}
  \begin{aligned}
    \text{impl} = \frac{1}{\lvert\mathbf{x} \in \mathbf{X}_{\mathbf{y}^*}\rvert} \sum_{\mathbf{x} \in \mathbf{X}_{\mathbf{y}^*}} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
  \end{aligned}
\end{equation}

where $\mathbf{X}_{\mathbf{y}^*}$ is a subsample of the training data in the target class $\mathbf{y}^*$. This gives rise to a very similar evaluation measure for unfaithfulness. We merely swap out the subsample of individuals in the target class for a subset $\hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^*}$ of the generated conditional samples:

\begin{equation}\label{eq:faith}
  \begin{aligned}
    \text{unfaith} = \frac{1}{\lvert\mathbf{x} \in \hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^*}\rvert} \sum_{\mathbf{x} \in \hat{\mathbf{X}}^{n_E}_{\theta,\mathbf{y}^*}} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
  \end{aligned}
\end{equation}

Specifically, we form this subset based on the $n_E$ generated samples associated with the lowest energy.

While we focus on these key evaluation metrics in the body of this paper, we also sporadically discuss outcomes with respect to other common measures used to evaluate the validity, proximity and sparsity of counterfactuals. Details can be found in Appendix~\ref{app:results}.

As noted by~\citet{guidotti2022counterfactual}, these distance-based measures are simplistic and more complex alternative measures may ultimately be more appropriate for the task. For example, we considered using statistical divergence measures instead. This would involve generating not one but many counterfactuals and comparing the generated empirical distribution to the target distributions in Definitions~\ref{def:plausible} and~\ref{def:faithful}. While this approach is potentially more rigorous, generating enough counterfactuals is not always practical.

@@

\begin{itemize}
  \item BatchNorm does not seem compatible with JEM
  \item Coverage and temperature impacts CCE in somewhat unpredictable ways
  \item It seems that models that are not explicitly trained for generative task, still learn it implictly
  \item Batch size seems to impact quality of generated samples (at inference, but not so much during JEM training)
  \item SGLD takes time 
  \item REVISE has benefit of lower dimensional space
  \item For MNIST it seems that ECCCo is better at reducing pixel values than increasing them (better at erasing than writing)
  \item JEMs are more difficult to train
  \item There is a tradeoff: higher cost vs. higher faithfulness/plausibility
  \item Results are sensitive to choices of penalty strength and step size
  \item Counterfactuals may end up looking fairly homogenous
  \item For MNIST data we found CP to have little effect
  \item JEMs themselves are sensitive to scale
  \item ECCCo can backfire, in case generative property of model is poor
\end{itemize}

@@

\subsection{Results}\label{app:results}

@@
